---
output: github_document
references:
- id: Touloumis2016
  title: "HDTD: Analyzing Multi-tissue Gene Expression Data"
  author:
  - family: Touloumis
    given: Anestis
  - family: Marioni 
    given: John C.
  - family: Tavaré
    given: Simon  
  container-title: Bioinfomatics
  volume: 32
  URL: 'https://doi.org/10.1093/bioinformatics/btw224'
  issue: 14
  page: 2193--2195
  type: article-journal
  issued:
    year: 2016
- id: Touloumis2015
  title: "Testing the Mean Matrix in High-Dimensional Transposable Data"
  author:
  - family: Touloumis
    given: Anestis
  - family: Tavaré
    given: Simon
  - family: Marioni 
    given: John C. 
  container-title: Biometrics
  volume: 71
  URL: 'http://onlinelibrary.wiley.com/doi/10.1111/biom.12054/full'
  issue: 1
  page: 157--166
  type: article-journal
  issued:
    year: 2015
- id: Touloumis2013
  title: "Hypothesis Testing for the Covariance Matrix in High-Dimensional Transposable Data with Kronecker Product Dependence Structure"
  author:
  - family: Touloumis
    given: Anestis
  - family: Marioni 
    given: John C.
  - family: Tavaré
    given: Simon  
  container-title: Submitted
  type: article-journal
  issued:
    year: 2017 
csl: Biometrics.csl  
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
  library(HDTD)
knitr::opts_chunk$set(
  tidy = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# HDTD: Analyzing High-Dimensional Transposable Data

[![Travis-CI Build Status](https://travis-ci.org/AnestisTouloumis/HDTD.svg?branch=master)](https://travis-ci.org/AnestisTouloumis/HDTD)
[![Project Status: Active The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active) 


## Installation

You can install the release version of `HDTD`:

``` r
## try http:// if https:// URLs are not supported
source("https://bioconductor.org/biocLite.R")
biocLite("HDTD")
```

The source code for the release version of `HDTD` is available on Bioconductor at:

- http://bioconductor.org/packages/release/bioc/html/HDTD.html

Or you can install the development version of `HDTD`:

``` r
# install.packages('devtools')
devtools::install_github("AnestisTouloumis/HDTD")
```

The source code for the development version of `HDTD` is available on Bioconductor at:

- http://bioconductor.org/packages/devel/bioc/html/HDTD.html

To use `HDTD`, you should load the package as follows:

``` r
library("HDTD")
```

## Usage

This package offers functions to estimate and test the matrix parameters of transposable data in high-dimensional settings. The term _transposable data_ refers to datasets that are structured in a matrix form such that both the rows and columns correspond to variables of interest. For example, consider microarray studies in genetics where multiple RNA samples across different tissues are available per subject. In this case, a data matrix can be created with row variables the genes, column variables the tissues and measurements the corresponding expression levels. For more examples see references in @Touloumis2013, @Touloumis2015 and @Touloumis2016.


There are four core functions:

- `meanmat.hat` to estimate the mean matrix of the transposable data,
- `meanmat.ts` to test the overall mean of the row (column) variables across groups of column (rows) variables,
- `covmat.hat` to estimate the row and column covariance matrix,
- `covmat.ts` to test the sphericity, identity and diagonality hypothesis for the row/column covariance matrix.


There are also three utility functions:

- `transposedata` for interchanching the role of rows and columns,
- `centerdata` for centering the transposable data around their mean matrix,
- `orderdata` for rearranging the order of the row and/or column variables.

## Example
We replicate the analysis that can be found in the vignette based on the mouse dataset 
```{r}
data(VEGFmouse)
```
One can estimate the mean relationship of the gene expression levels across the $9$ tissues
```{r}
sample_mean <- meanmat.hat(datamat = VEGFmouse,N=40)
sample_mean
```
and test whether the overall gene expression is constant across the $9$ tissues:
```{r}
tissue_test <- meanmat.ts(datamat = VEGFmouse,N=40,group.sizes=9)
tissue_test
```
In this case, the overall gene expression is not conserved. 

To analyze the gene-wise and tissue-wise dependence structure, one needs to estimate the two covariance matrices:
```{r}
est_cov_mat <- covmat.hat(datamat=VEGFmouse,N=40)
est_cov_mat
```
Finally, the package allows users to perform hypothesis tests for the covariance matrix of the genes
```{r}
genes_cov_test <- covmat.ts(VEGFmouse,N=40)
genes_cov_test
```
and of the tissues:
```{r}
tissues_cov_test <- covmat.ts(VEGFmouse,N=40,voi="columns")
tissues_cov_test
```
At a $5\%$ significance level, it appears that the genes are correlated but we do not have enough evidence to reject the hypothesis that the tissues are uncorrelated. 

## Getting help
The statistical methods implemented in `HDTD` are described in @Touloumis2013, @Touloumis2015 and @Touloumis2016. Detailed examples of `HDTD` can be found in @Touloumis2016 or in the vignette: 

```{r, eval=FALSE}
browseVignettes("HDTD")
```


## How to cite

```{r, echo=FALSE, comment=""}
print(citation("HDTD"),bibtex=TRUE)
```

# References